---
title: "DPEA"
author: "Sophie.Alice"
date: "2026-07-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---


# Setup

```{r, message=FALSE, warning = FALSE}
rm(list = ls())
gc()

library(here)
library(dplyr)
library(readr)
library(stringr)
library(tibble)
library(ggplot2)
library(OlinkAnalyze)
library(DT)
library(limma)
library(tidyverse)
library(tidyr)
library(ggrepel)
library(clusterProfiler)
library(org.Hs.eg.db)
library(fgsea)
library(purrr)
library(aPEAR)

```

# Functions

```{r, warning = FALSE}


##################################### Tables

show_table <- function(df) {
  DT::datatable(
    df,
    extensions = 'Buttons', 
    options = list(
      dom = 'Blfrtip',                       
      buttons = c('copy', 'csv', 'excel', 'pdf'),
      pageLength = 10,                      
      lengthMenu = c(10, 25, 50, 100),
      autoWidth = TRUE,
      columnDefs = list(list(
        targets = '_all',
        className = 'dt-nowrap'
      ))
    ),
    filter = 'top',                         
    class = 'cell-border stripe hover nowrap'       
  )
}


```

# Import Data

```{r, warning = FALSE}

data_path <- "/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Olink_Project/OLINK_Data/RDSfiles/filtered_NPXs_annot.rds"

filt.NPX <- readRDS(data_path)

custom_palette <- colorRampPalette(c("#005EA8", "#6DB335", "#FFD966", "#EDAAD2", "#9F1C67"))
ann_colors <- custom_palette(8)
colors5 <- custom_palette(5)
```

# Check inter Group differences with ANOVA

```{r, warning = FALSE, fig.width=8, fig.height=6}
# Run ANOVA on NPX across Experimental Groups
anova_results <- olink_anova(
  df = filt.NPX,
  variable = "Group"
)

sig_anova_ids <- anova_results |>
  dplyr::filter(Threshold == "Significant") |>
  dplyr::pull(OlinkID)

# Heatmap showing global group clustering
olink_heatmap_plot(
  df = filt.NPX,
  variable_row_list = c("Group", "Experiment"),
  olinkid_list = sig_anova_ids,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,   
  show_colnames = FALSE,
  annotation_colors = ann_colors
)

length(sig_anova_ids)

# -> 260 proteins are differently expressed between the groups
```


# DPEA with limma

```{r, warning = FALSE}

# NPX data into matrix format (Proteins = Rows, Samples = Columns)
npx_mat <- filt.NPX |>
  dplyr::select(SampleID, Assay, NPX) |>
  tidyr::pivot_wider(names_from = SampleID, values_from = NPX, values_fn = mean) |>
  tibble::column_to_rownames("Assay") |>
  as.matrix()

# Extract sample metadata matching the matrix column order
sample_meta <- filt.NPX |>
  dplyr::select(SampleID, Group, Experiment) |>
  dplyr::distinct() |>
  dplyr::slice(match(colnames(npx_mat), SampleID))

# Clean group names for limma syntax (removes spaces and special characters)
clean_group <- factor(make.names(sample_meta$Group))

# Create Design Matrix
design <- model.matrix(~ 0 + clean_group)
colnames(design) <- levels(clean_group)


#############################################

# Fit initial linear model across all proteins
fit <- lmFit(npx_mat, design)

# Define contrasts
contrast_matrix <- makeContrasts(
  # 1. Macrophage effect under TGFb
  HM_TGF_vs_NC_TGF = MT...Macrophages..TGFb. - MT.Control..TGFb.,
  # 2. TGFb effect in MT without macrophages
  NC_TGF_vs_NC_NC  = MT.Control..TGFb. - MT.Control..NC.,
  # 3. Macrophage effect at baseline (NC)
  HM_NC_vs_NC_NC   = MT...Macrophages..NC. - MT.Control..NC.,
  # 4. TGFb effect with Macrophages present
  HM_TGF_vs_HM_NC  = MT...Macrophages..TGFb. - MT...Macrophages..NC.,
  # 5. The Interaction term = How does Mac presence affect the response to TGFb?
  Interaction      = (MT...Macrophages..TGFb. - MT...Macrophages..NC.) - 
                     (MT.Control..TGFb. - MT.Control..NC.),
  levels = design
)

# Fit contrasts & apply Empirical Bayes moderation
fit_contrasts <- contrasts.fit(fit, contrast_matrix)
fit_ebayes    <- dBayes <- eBayes(fit_contrasts)

#############################################


# Create empty list to hold tables
limma_results_list <- list()

# Loop through all 5 contrast names and store tables
for (c_name in colnames(contrast_matrix)) {
  
  res_table <- topTable(fit_ebayes, coef = c_name, number = Inf, adjust.method = "BH") |>
    tibble::rownames_to_column("OlinkID") |>
    dplyr::mutate(
      Significance = dplyr::case_when(
        adj.P.Val < 0.05 & logFC > 0.5  ~ "Up",
        adj.P.Val < 0.05 & logFC < -0.5 ~ "Down",
        TRUE                           ~ "Not Sig"
      )
    )
  
  # Add to named list
  limma_results_list[[c_name]] <- res_table
}

```

# Showing the DEP Tables

```{r, warning = FALSE}
show_table(limma_results_list$NC_TGF_vs_NC_NC)
show_table(limma_results_list$HM_TGF_vs_HM_NC)
show_table(limma_results_list$HM_NC_vs_NC_NC)
show_table(limma_results_list$HM_TGF_vs_NC_TGF)
show_table(limma_results_list$Interaction)
```

# showing the DEPs with Volcanoplots

```{r, warning = FALSE, fig.width=8, fig.height=6}
# Loop through the list of tables and render volcano plots
for (c_name in names(limma_results_list)) {
  
  # get data frame for current comparison
  res_table <- limma_results_list[[c_name]]
  
  # Count up/down significant hits
  n_up <- sum(res_table$Significance == "Up")
  n_down <- sum(res_table$Significance == "Down")
  
  # Volcano Plot
  p <- ggplot(res_table, aes(x = logFC, y = -log10(adj.P.Val), color = Significance)) +
    geom_point(alpha = 0.8, size = 2) +
    geom_text_repel(
      data = res_table |> dplyr::filter(Significance != "Not Sig") |> dplyr::slice_min(adj.P.Val, n = 15),
      aes(label = OlinkID),
      size = 3.5,
      max.overlaps = 15,
      show.legend = FALSE
    ) +
    scale_color_manual(values = c("Up" = "#9F1C67", "Down" = "#005EA8", "Not Sig" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
    geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted", color = "black") +
    theme_bw(base_size = 12) +
    labs(
      title = paste("DEPs:", c_name),
      subtitle = paste0("Up: ", n_up, " | Down: ", n_down, " (FDR < 0.05, |logFC| > 0.5)"),
      x = "Log2 Fold Change (NPX Difference)",
      y = "-log10 FDR"
    ) +
    theme(plot.title = element_text(face = "bold", size = 12),
          plot.subtitle = element_text(face = "italic", size = 10))
  
  print(p)
}
```

# Deeper look into the interaction term Protein hits

Aka How does Mac presence affect the response to TGFb?

Genes that have a > 0.5 log2FC amplify the TGFb response
Genes that have a < -0.5 log2FC dampen the TGFb response

```{r, warning = FALSE}
# Extract top up proteins (FDR < 0.05 and logFC > 0.5) -> TOP 6
up_hits <- limma_results_list$Interaction |>
  dplyr::filter(adj.P.Val < 0.05 & logFC > 0.5) |>
  dplyr::arrange(adj.P.Val)

# Extract top up proteins (FDR < 0.05 and logFC < 0.5) -> TOP 6
down_hits <- limma_results_list$Interaction |>
  dplyr::filter(adj.P.Val < 0.05 & logFC < -0.5) |>
  dplyr::arrange(adj.P.Val)

top_up_assays   <- up_hits$OlinkID
top_down_assays <- down_hits$OlinkID

################################## show boxplots up hits

# Define chunk size 
chunk_size <- 8
assay_chunks <- split(top_up_assays, ceiling(seq_along(top_up_assays) / chunk_size))


for (i in seq_along(assay_chunks)) {
  
  current_assays <- assay_chunks[[i]]
  
  p <- filt.NPX |>
    dplyr::filter(Assay %in% current_assays) |>
    dplyr::mutate(Assay = factor(Assay, levels = current_assays)) |>
    ggplot(aes(x = Group, y = NPX)) +
    geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.6) +
    geom_jitter(
      aes(color = AboveLOD), 
      width = 0.15, size = 2, alpha = 0.85
    ) +
    scale_fill_manual(values = colors5) +
    
    scale_color_manual(
      values = c("FALSE" = "darkred", "TRUE" = "black"),
      labels = c("Below LOD", "Above LOD"),
      name = "Detection") +
    facet_wrap(~ Assay, scales = "free_y", ncol = 4) +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom",
      strip.background = element_rect(fill = "darkgrey", color = NA),
      strip.text = element_text(color = "black", face = "bold"),
      plot.title = element_text(face = "bold", size = 12),
      plot.subtitle = element_text(face = "italic", size = 10)) +
    labs(
      title = paste0("Amplifying Hits (Page ", i, " of ", length(assay_chunks), ")"),
      subtitle = "Red dots = hits below the LOD",
      x = NULL,
      y = "NPX Value"
    )
  print(p)
}

################################## show boxplots down hits

# Define chunk size 
chunk_size <- 8
assay_chunks <- split(top_down_assays, ceiling(seq_along(top_down_assays) / chunk_size))


for (i in seq_along(assay_chunks)) {
  
  current_assays <- assay_chunks[[i]]
  
  p <- filt.NPX |>
    dplyr::filter(Assay %in% current_assays) |>
    dplyr::mutate(Assay = factor(Assay, levels = current_assays)) |>
    ggplot(aes(x = Group, y = NPX)) +
    geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.6) +
    geom_jitter(
      aes(color = AboveLOD), 
      width = 0.15, size = 2, alpha = 0.85
    ) +
    scale_fill_manual(values = colors5) +
    
    scale_color_manual(
      values = c("FALSE" = "darkred", "TRUE" = "black"),
      labels = c("Below LOD", "Above LOD"),
      name = "Detection") +
    facet_wrap(~ Assay, scales = "free_y", ncol = 4) +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom",
      strip.background = element_rect(fill = "darkgrey", color = NA),
      strip.text = element_text(color = "black", face = "bold"),
      plot.title = element_text(face = "bold", size = 12),
      plot.subtitle = element_text(face = "italic", size = 10)) +
    labs(
      title = paste0("Dampening Hits (Page ", i, " of ", length(assay_chunks), ")"),
      subtitle = "Red dots = hits below the LOD",
      x = NULL,
      y = "NPX Value"
    )
  print(p)
}

```

# fGSEA with Hallmark pathways

```{r, warning = FALSE}

Hallmark.pathways <- gmtPathways("/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/Hallmark/h.all.v2026.1.Hs.symbols.gmt")

##################################### Function of Hallmark fGSEA

run_gsea_pipeline <- function(degs_df, comparison_name, pathways = Hallmark.pathways, top_n = 15) {
  
  # Clean data & create ranked named vector
  protein_ranked_df <- degs_df %>%
    dplyr::select(OlinkID, logFC) %>% 
    na.omit() %>%
    distinct() %>%
    group_by(OlinkID) %>%
    summarize(log2FC = mean(logFC, na.rm = TRUE)) %>%
    arrange(desc(log2FC))
  
  ranked_proteins <- deframe(protein_ranked_df)
  
  # Run GSEA (Using fgseaMultilevel)
  fgseares <- fgseaMultilevel(pathways = pathways, stats = ranked_proteins)
  
  # Process Tidy Results
  fgseares.tidy <- fgseares %>%
    as_tibble() %>%  
    arrange(desc(NES))
  
  # Extract Top Up and Down pathways
  fgseares.tidy.top.up   <- head(fgseares.tidy, top_n)
  fgseares.tidy.top.down <- tail(fgseares.tidy, top_n)
  fgseares.tidy.top.both <- bind_rows(fgseares.tidy.top.up, fgseares.tidy.top.down)
  
  # Format dataframe for plotting
  plot_df <- fgseares.tidy.top.both %>%
    mutate(
      padj_group = case_when(
        padj <= 0.05 ~ "< 0.05",
        padj <= 0.2  ~ "0.05 – 0.2",
        TRUE         ~ "> 0.2"
      ),
      padj_group = factor(padj_group, levels = c("< 0.05", "0.05 – 0.2", "> 0.2")),
      pathway = gsub("HALLMARK_", "", pathway)
    )
  
  # Generate Plot
  p <- ggplot(plot_df, aes(reorder(pathway, NES), NES)) +
    geom_col(aes(fill = padj_group), color = "black", linewidth = 0.2) +
    coord_flip() +
    scale_x_discrete(position = "top", labels = scales::label_wrap(50)) +
    scale_fill_manual(
      name = "adj. p-value",
      values = c(
        "< 0.05"     = "#9F1C67",   
        "0.05 – 0.2" = "#FFD966",   
        "> 0.2"      = "#005EA8"    
      ),
      drop = FALSE 
    ) +
    labs(
      x = "", 
      y = "Normalized Enrichment Score (NES)", 
      title = "GSEA: Hallmark Pathways",
      subtitle = paste("Comparison:", comparison_name)
    ) +
    theme_bw() +
    theme(
      aspect.ratio    = 1.4, 
      axis.text.y     = element_text(size = 8, hjust = 0), 
      axis.title      = element_text(face = "bold"),
      plot.title      = element_text(hjust = 0, face = "bold"),
      plot.subtitle   = element_text(hjust = 0, size = 9, face = "italic"),
      legend.position = "left" 
    )
  
  # Return both table and plot in a list
  return(list(
    results_table = fgseares.tidy,
    plot          = p
  ))
}

########################################

# Run GSEA for every dataset in the list
gsea_results_list <- imap(limma_results_list, function(df, contrast_name) {
  run_gsea_pipeline(degs_df = df, comparison_name = contrast_name)
})

# show results
gsea_results_list$HM_TGF_vs_NC_TGF$plot
show_table(gsea_results_list$HM_TGF_vs_NC_TGF$results_table)

gsea_results_list$NC_TGF_vs_NC_NC$plot
show_table(gsea_results_list$NC_TGF_vs_NC_NC$results_table)

gsea_results_list$HM_NC_vs_NC_NC$plot
show_table(gsea_results_list$HM_NC_vs_NC_NC$results_table)

gsea_results_list$HM_TGF_vs_HM_NC$plot
show_table(gsea_results_list$HM_TGF_vs_HM_NC$results_table)

gsea_results_list$Interaction$plot
show_table(gsea_results_list$Interaction$results_table)


```


# fGSEA with ClusterProfiler 

```{r, warning = FALSE}

protein_ranked_df <- limma_results_list$HM_TGF_vs_HM_NC %>%
    dplyr::select(OlinkID, logFC) %>% 
    na.omit() %>%
    distinct() %>%
    group_by(OlinkID) %>%
    summarize(log2FC = mean(logFC, na.rm = TRUE)) %>%
    arrange(desc(log2FC))
  
ranked_proteins <- deframe(protein_ranked_df)
  

gse_go_res <- gseGO(
  geneList     = ranked_proteins,
  OrgDb        = org.Hs.eg.db,
  keyType      = "SYMBOL",       
  ont          = "BP",           
  minGSSize    = 10,
  maxGSSize    = 500,
  pvalueCutoff = 0.1,
  pAdjustMethod = "BH"
)

dotplot(
  gse_go_res, 
  showCategory = 10, 
  split = ".sign", 
  title = "GO Biological Processes: Activated vs Suppressed"
) + 
  facet_grid(. ~ .sign) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 9),
    plot.title  = element_text(face = "bold", size = 12)
  )

# 
# gsea_sig_df <- gse_go_res@result %>%
#   dplyr::filter(p.adjust < 0.05)  
# 
# p_apear <- enrichmentNetwork(
#   enrichment = gsea_sig_df,
#   repelLabels = TRUE,
#   drawEllipses = TRUE,
#   fontSize     = 3,
#   simMethod = "jaccard",
#   clustMethod = "hier",
#   colorBy      = "NES",        
#   colorType    = "nes"  
# )
# 
# p_apear <- p_apear + labs(
#   title    = "Network of Significant GO-BP Terms Identified by GSEA",
#   subtitle = "MT-Fibroblasts: differences in CMF_TGF compared to CMF_NC",
#   color    = "NES",       
# ) +
# theme(
#   plot.title.position = "plot",
#   plot.title    = element_text(face = "bold", size = 12, hjust = 0),
#   plot.subtitle = element_text(face = "italic", size = 10, hjust = 0)
# )
# 
# print(p_apear)

```


# ORA

## GO PB

 niet 
 
```{r, warning = FALSE}

up_hits <- limma_results_list$HM_TGF_vs_NC_TGF |>
  dplyr::filter(adj.P.Val < 0.05 & logFC > 0.5) |>
  dplyr::arrange(adj.P.Val)

# Extract top up proteins (FDR < 0.05 and logFC < 0.5) -> TOP 6
down_hits <- limma_results_list$HM_TGF_vs_NC_TGF |>
  dplyr::filter(adj.P.Val < 0.05 & logFC < -0.5) |>
  dplyr::arrange(adj.P.Val)

top_up_assays   <- up_hits$OlinkID
top_down_assays <- down_hits$OlinkID

# GO Biological Process Enrichment
ego_amp <- enrichGO(
  gene          = top_up_assays,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP", # Biological Process
  pAdjustMethod = "none",
  pvalueCutoff  = 0.9
)

# Visualize Top 10 Enriched Terms
#dotplot(ego_amp, showCategory = 10, title = "GO BP: Amplifying Hits")



ego_amp <- enrichGO(
  gene          = top_down_assays,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP", # Biological Process
  pAdjustMethod = "none",
  pvalueCutoff  = 0.5
)

# Visualize Top 10 Enriched Terms
#dotplot(ego_amp, showCategory = 10, title = "GO BP: Amplifying Hits")
```



```{r, warning = FALSE}


```

```{r, warning = FALSE}


```

```{r, warning = FALSE}

```

```{r, warning = FALSE}

```

```{r, warning = FALSE}

```

